Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-08-28 RNARedPrint-Bioinfo2018.tex
Work on comments of reviewer R1---leave some open todo notes

%\usepackage{todonotes}

@@

\newcommand{\Def}[1]{{\bfseries #1}}

@@

\enlargethispage{35.1pt}
\section{Introduction}
\parHead{Design, applications and motivation for multiple design.}Synthetic biology endeavors the engineering of artificial biological
systems, promising broad applications in biology, biotechnology and
medicine. Centrally, this requires the design of biological
macromolecules with highly specific properties and programmable functions.
RNA present itself as a well-suited tool for a
rational design targeting specific functions~\citep{Kushwaha2016}. RNA function is tightly
coupled to the formation of secondary structure, as well as changes in
base pairing propensities and the accessibility of regions, \emph{eg} by
burying or exposing interaction sites~\citep{Rodrigo2014}. At the same time, the
thermodynamics of RNA secondary structure is well understood and its prediction is
computationally tractable~\citep{McCaskill1990}. Thus,  structure can serve as effective
proxy for rational design approaches, ultimately targeting catalytic~\citep{Zhang2013} or regulatory functions.

@@

\parHead{On the importance of sampling for design.}
Many computational methods for RNA design rely on similar overall
strategies: initially generating one or several \Def{seed} sequences
and optimizing them subsequently. In many cases, the seed quality was
found to be critical for the empirical performance of RNA design
methods~\citep{Levin2012}. For instance, random seed generation
improves the prospect of subsequent optimizations, helping to overcome
local optima of the objective function, and increases the diversity
across designs~\citep{Reinharz2013}.  For single-target approaches,
\Software{INFO-RNA}~\citep{Busch2006} made significant improvements
mainly by starting its local search from the minimum energy sequence
for the target structure instead of (uniform) random sequences for the
early \Software{RNAinverse} algorithm~\citep{Hofacker1994}. This
strategy was later shown to result in unrealistically high
\GCb\% in designed sequences. To address this issue,
\Software{IncaRNAtion}~\citep{Reinharz2013} controls \GCb\%
through an adaptive sampling strategy.

\parHead{Specificities and similarities of multi-target design.}
Specifically, for multi-target design, virtually all available methods~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generation/optimization scheme.
%
Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} did not attempt to solve sequence generation systematically, but rely on \emph{ad-hoc} sampling strategies.
%
Recently, the \Software{RNAdesign}
approach~\citep{HoenerzuSiederdissen2013}, coupled with powerful local
search within \RNAblueprint{}~\citep{Hammer2017}, solved the
problem of sampling seeds from the uniform distribution for multiple
targets. These methods adopt a graph coloring perspective, initially
decomposing the graph hierarchically using various decomposition
algorithms, and \Def{precomputing} the number of valid sequences
within each subgraph. The decomposition is then reinterpreted as a
decision tree to perform a \Def{stochastic backtrack}, inspired by
Ding and Lawrence~\citep{Ding2003}. Uniform sampling is achieved by
choosing individual nucleotide assignments with probabilities derived
from the subsolution counts. The overall complexity of
\Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where the
parameter $\gamma$ is bounded by the length of the designed RNA;
typically, the decomposition strategy achieves much lower $\gamma$.

@@

To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures (Section~\ref{sec:FPT}). Guided by a \Def{tree decomposition} of the network, we devise dynamic programming to compute partition functions and sample sequences from the Boltzmann distribution%
% (Subsection~\ref{sec:PF})
. We show that these algorithms are \Def{fixed-parameter tractable} for the \Def{treewidth}%
% Subsection~\ref{sec:complexity}
%
; in practice, we limit this parameter by using state-of-the-art tree decomposition algorithms.
%Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
By evaluating (partial) sequences in a weighted constraint
network, we support arbitrary multi-ary constraints and thus
arbitrarily complex energy models,
notably subsuming all commonly
used RNA energy models%
%(Subsection~\ref{sec:energy_models})
.  Moreover, we describe an \Def{adaptive
  sampling} strategy to control the free energies of the individual
target structures and \GCb\%%
% (SubSection~\ref{sec:multiBoltzmann})
. %
We observe that sampling based on less complex RNA energy models
(taking only the most important energy contributions into account)
still allows targeting realistic RNA energies in the well-accepted
Turner RNA energy model. The resulting combination of efficiency and
high accuracy finally enables generating biologically relevant
multi-target designs in our final application of our overall strategy
to a large set of multi-target RNA design instances from a
representative benchmark Section~\ref{sec:results}).

@@

%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
%
  As we elaborate in subsequent sections, our approach relies on
  breaking down the energy function $\Energy{\R}{S}$ into additive
  components, each depending on only few sequence positions. Given
  $\R$, we express $\Energy{\R}{S}$ as the sum of energy contributions $f(S)$
  over a set $\F$ (of functions $f:\S_n\to\real$),
  s.t.~$\Energy{\R}{S}=\sum_{f\in\F} f(S)$. This captures realistic RNA
  energy models---including nearest neighbor models,
  e.g.~\citet{Turner2009}, and even pseudoknot models,
  e.g. \citet{Andronescu2010}, while bounding the dependencies to
  sequence positions introduced by each single $f$. Formally, define
  the \Def{dependencies} $\dep(f)$ of $f$ as the minimum set of sequence
  positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all
  sequences $S$ and $S'$ that agree at the positions in $\I$.

  Each set $\F$ of functions (on sequences of length $n$) induces a
  \Def{dependency graph} on sequence positions, namely the hypergraph
  $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$.  Our algorithms
  will critically rely on a \Def{tree decomposition} of the dependency
  graph, which we define below.
\begin{definition}[Tree decomposition and width]
  \label{def:treedecomp}
  Let $G=(X, E)$ be a hypergraph. A \Def{tree decomposition} of $G$ is
  a pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and (for
  each $v\in T$) $\chi(v)\subseteq X$ is a set of vertices assigned to
  the node tree $v\in T$, such that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item for all $x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
\item for all $e\in E$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}
The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
$\width(T,\chi) = \min_{u\in T} |\chi(u)| - 1 $. The \Def{treewidth}
of $G$ is the smallest width of any tree decomposition of $G$.
\end{definition}

\section{An FPT algorithm for the partition function and sampling of Boltzmann-weighted designs}
\label{sec:FPT}

@@

\subsection{Partition function and Boltzmann sampling through stochastic backtrack}\label{sec:PF}
The minimum energy, counting, and partition function
over RNA design network can be computed by dynamic programming based
on a tree decomposition of the network's dependency graph
(i.e. cluster tree elimination).
%This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.
We focus on the efficient computation of the partition
function. %Observe that counting is a special case of the partition function and can be obtained by setting $\beta =0$.%; moreover, efficient energy minimization can be obtained by a simple $(+,\times)\to (\min,+)$ algebraic substitution and dropping the transformation of energy contributions to Boltzmann weights.

@@

For a set $\Y$ of sequence positions, write $\partseqs(\Y)$ to
denote the set of all partial sequences that determine exactly the positions
in $\Y$; furthermore, given partial sequences $\val$ and
$\val'$, we define the \Def{combined partial sequence $\substitute{\val'}{\val''}$} such that
$$
(\substitute{\val}{\val'})_i :=
\begin{cases}
  \val'_i & \text{if } \val'_i\in \Sigma\\
  \val_i & \text{otherwise}
\end{cases}
$$

@@

Let us now consider the \Def{computation of the partition function}.
Given is the RNA design network $\network=(\X,\F)$ and its cluster
tree decomposition $(T,\chi,\phi)$.  W.l.o.g., we assume that $T$ is
connected and contains a dedicated node $r$, with
$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$, added as a virtual
root $r$ connected to a node in each connected component of $T$.  Now,
we consider the set of directed edges $\edgesToR{}$ of $T$ oriented to
$r$; define $T_r(u)$ as the induced subtree of
$u$. Algorithm~\ref{alg:pf} computes the partition function by passing
messages along these directed edges $u\to v$ (i.e. always from some
child $u$ to its parent $v$). Each message is a function that depends
on the positions $\dep(m)\subseteq \X$ and yields a partition function
in $\real\cup\{\infty\}$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Supp. Mat.~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function:
\begin{math}
  \partfun{\network} = \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

\begin{algorithm}[t]
  \KwData{Cluster tree $(T,\chi,\phi)$} \KwResult{Messages
    $\Message{u}{v}$ for all $(u\to{}v)\in T$; i.e.~partition
    functions of the subtrees of all $v$ for all possible partial
    sequences determining exactly the positions $\separator{u}{v}$.}
 \For{$u\to{}v\in T$ in postorder}{
  \For{$\val\in\partseqs(\separator{u}{v})$}{
    $x\gets 0$\;
    \For{$\val'\in\partseqs(\difference{u}{v})$}{
     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
     ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
     $x \gets x + p$\;
   }
   $\evalfor{\Message{u}{v}}{\val} \gets x$\;
  }
  \Return {$m$}\;
  }
 \caption{FPT computation of the partition function using
   dynamic programming (CTE). }\label{alg:pf}
\end{algorithm}

\SetKwProg{Fn}{Function}{}{}
 \SetKwFunction{Sample}{$\sample$}
 \SetKwFunction{Random}{UnifRand}

The partition functions can then direct a \Def{stochastic backtrack} to achieve \Def{Boltzmann sampling of sequences}, such that one samples from the Boltzmann distribution of a given design network $\network$. The sampling algorithm assumes that the cluster tree was expanded and the messages $\Message{u}{v}$ for the edges in $\edgesToR{}$ are already generated by Algorithm \ref{alg:pf} for the expanded cluster tree.
%
Algorithm~\ref{alg:sampling} defines the recursive procedure
$\sample(u,\val)$, which returns---randomly drawn from the
Boltzmann distribution---a partial sequence that determines all
sequence positions in the subtree rooted at $u$.  Called on $r$ and
the empty partial sequence, which does not determine any positions,
the procedure samples a random sequence from the Boltzmann
distribution.

@@

We define the \emph{maximum separator size} $s$ as
$\max_{u,v\in V} | \separator{u}{v} |$ and denote by  $D$ the maximum size of
$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$.  In the absence
of specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and
$\mathcal{O}(|V|\cdot4^s)$ space
(Supp. Mat.~\ref{appsec:algcomplexity}); Alg.~\ref{alg:sampling} would
require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per sample on arbitrary
tree decompositions
(Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g. we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

\begin{algorithm}
 \KwData{Node $u$, partial sequence $\val\in\partseqs(\separator{u}{v})$;\newline
 Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to{}v')\in T$ and $\val'\in\partseqs(\separator{u'}{v'})$.}
 \KwResult{Boltzmann-distributed random partial sequence for the subtree rooted at $u$, specializing a partial sequence $\val$.}
 \Fn{\Sample$(u,\val;T,\chi,\phi,m)$}{
   $r \gets \Random(\evalfor{\Message{u}{v}}{\val})$\;
   \For{$\val'\in\partseqs(\difference{u}{v})$}{

     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$\\
     ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{w}{u}}{\substitute{\val}{\val'}}$ for $(w\to{}u)\in T$ )\;
%
%     $p \gets 1$\;
%     \For{$f\in \phi(v)$}{
%  	      $p \gets p \cdot \exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$\;
%     }
%     \For{$v\to{}w\in T$}{
%  	      $p \gets p \cdot \evalfor{\Message{v}{w}}{\substitute{\val}{\val'}}$\;
%     }
  	  $r \gets r - p$\;
  	  \If{$r<0$}{
  	    $\val_{\rm res} \gets \substitute{\val}{\val'}$\;
  	  \For{$(v\to{}w)\in T$}{
  	      $\val_{\rm res} \gets \substitute{\val_{\rm res}}{\Sample(v,\substitute{\val}{\val'};T,\chi,\phi,m)}$\;
     }
  	  \Return {$\val_{\rm res}$}
  	  \;}
 }
 }
 \caption{Stochastic backtrack algorithm for partial sequences in the Boltzmann distribution.}\label{alg:sampling}
\end{algorithm}
%For example, for design to a single non-crossing target in the nearest neighbor energy model, the complexity of Algs.~\ref{alg:pf} and \ref{alg:sampling} degrades to linear time (as reported for IncaRNAtion), since there are only $O(n)$ many functions and nodes (in the sequence length $n$). Moreover, the dependencies are tree-like owing to the tree-like non-crossing structure; this implies a constantly bounded maximum treewidth.

\begin{theorem}[Complexities]\label{th:complexities}
  For sequence length $n$, $k$ target structures, treewidth $w$ and
  a base pair dependency graph having $c$ connected components, $t$ sequences
  are generated from the Boltzmann distribution  in
  $O( n\, k \, 4^{w+1} + t\, n\, k )$ time.
  %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
  %$\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
\end{theorem}
As shown in Supp. Mat.~\ref{sec:improvedComplexity}, the complexity of the precomputation can be further improved to
$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ is the maximum number of connected components represented in a node of the tree decomposition ($c\le w+1$).

\subsection{Sequence features, constraints, and energy
  models.}\label{sec:energy_models}

The functions in $\F$ allow expressing complex features of the
sequences alone, e.g. rewarding or penalizing specific sequence
motifs, as well as features depending on the target structures.
Furthermore, constraints, which enforce or forbid features, are
naturally expressed by assigning infinite penalties to invalid
(partial) sequences. Finally, but less obviously, the framework
captures various RNA energy models with bounded dependencies, which we
describe briefly.

% To illustrate the expressivity of the framework, we outline the construction of function sets that express two different energy models for the energies $E(S;R_\ell)$ for encoding
% \begin{equation}
%   \label{eq:nussinov-network-energy}
%   \energy{\network}(S) = \sum_{\ell\in[1,k]} w_\ell E(S;R_\ell)
% \end{equation}
In simple \Def{base pair-based energy models}, energy is defined as
the sum of base pair (pseudo-)energies. If base pair energies
$\Ebp{k}{i,j,x,y}$ (where $i$ and $j$ are sequence positions, $x$ and
$y$ are bases in $\Sigma$) are given for each target structure $\ell$,
s.t.  $ E(S;R_\ell) := \sum_{(i,j)\in R_\ell} \Ebp{k}{i,j,S_i,S_j}$,
we encode the network energy by the set of functions $f$ for each base
pair $(i,j)\in R_\ell$ of each input structure $R_\ell$ that evaluate
to $\ln(\pi_\ell) \Ebp{k}{i,j,\val(x_i),\val(x_j)}$ under partial sequence $\val$;
here, $\pi_\ell>0$ is a weight that controls the
influence of structure $\ell$ on the sampling (as elaborated later).

>>>>

\usepackage{todonotes}

@@

\newcommand{\Def}[1]{\emph{#1}}

@@

\enlargethispage{35.1pt}
\section{Introduction}
\parHead{Design, applications and motivation for multiple design.}Synthetic biology endeavors the engineering of artificial biological
systems, promising broad applications in biology, biotechnology and
medicine. Centrally, this requires the design of biological
macromolecules with highly specific properties and programmable functions.
RNA present itself as a well-suited tool for a
rational design targeting specific functions~\citep{Kushwaha2016}. RNA function is tightly
coupled to the formation of secondary structure, as well as changes in
base pairing propensities and the accessibility of regions, e.g.\ by
burying or exposing interaction sites~\citep{Rodrigo2014}. At the same time, the
thermodynamics of RNA secondary structure is well understood and its prediction is
computationally tractable~\citep{McCaskill1990}. Thus,  structure can serve as effective
proxy for rational design approaches, ultimately targeting catalytic~\citep{Zhang2013} or regulatory functions.

@@

\parHead{On the importance of sampling for design.}
Many computational methods for RNA design rely on similar overall
strategies: initially generating one or several \Def{seed} sequences
and optimizing them subsequently. In many cases, the seed quality was
found to be critical for the empirical performance of RNA design
methods~\citep{Levin2012}. For instance, random seed generation
improves the prospect of subsequent optimizations, helping to overcome
local optima of the objective function, and increases the diversity
across designs~\citep{Reinharz2013}.  For single-target approaches,
\Software{INFO-RNA}~\citep{Busch2006} made significant improvements
mainly by starting its local search from the minimum energy sequence
for the target structure instead of (uniform) random sequences for the
early \Software{RNAinverse} algorithm~\citep{Hofacker1994}. This
strategy was later shown to result in unrealistically high
\GCb\% in designed sequences. To address this issue,
tools like \Software{antaRNA}~\citep{Kleinkauf2015} and \Software{IncaRNAtion}~\citep{Reinharz2013} control \GCb\%; the latter based on adaptive sampling.

\parHead{Specificities and similarities of multi-target design.}
Specifically, for multi-target design, virtually all available methods~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} follow the same overall generate-and-optimize strategy.
%
Facing the complex sequence constraints induced by multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} did not attempt to solve sequence generation systematically, but rely on ad-hoc sampling strategies.
%
Recently, the \Software{RNAdesign}
approach~\citep{HoenerzuSiederdissen2013}, coupled with powerful local
search within \RNAblueprint{}~\citep{Hammer2017}, solved the
problem of sampling seeds from the uniform distribution for multiple
targets. These methods adopt a graph coloring perspective, initially
decomposing the graph hierarchically using various decomposition
algorithms, and \Def{precomputing} the number of valid sequences
within each subgraph. The decomposition is then reinterpreted as a
decision tree to perform a \Def{stochastic backtrack}, inspired by
Ding and Lawrence~\citep{Ding2003}. Uniform sampling is achieved by
choosing individual nucleotide assignments with probabilities derived
from the subsolution counts. The overall complexity of
\Software{RNAdesign} grows like $\Theta(4^{\gamma})$, where the
parameter $\gamma$ is bounded by the length of the designed RNA;
typically, the decomposition strategy achieves much lower $\gamma$.

@@

To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures (Section~\ref{sec:FPT}). Guided by a \Def{tree decomposition}, we devise dynamic programming to compute partition functions and sample sequences from the Boltzmann distribution%
% (Subsection~\ref{sec:PF})
. We show that these algorithms are \Def{fixed-parameter tractable} for the \Def{treewidth}%
% Subsection~\ref{sec:complexity}
%
; in practice, we limit this parameter by using state-of-the-art tree decomposition algorithms.
%Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
By evaluating (partial) sequences in a weighted constraint
network, we support arbitrary multi-ary constraints and thus
arbitrarily complex energy models,
notably subsuming all commonly
used RNA energy models%
%(Subsection~\ref{sec:energy_models})
.  Moreover, we describe an \Def{adaptive
  sampling} strategy to control the free energies of the individual
target structures and \GCb\%%
% (SubSection~\ref{sec:multiBoltzmann})
. %
We observe that sampling based on less complex RNA energy models
(taking only the most important energy contributions into account)
still allows targeting realistic RNA energies in the well-accepted
Turner RNA energy model. The resulting combination of efficiency and
high accuracy finally enables generating biologically relevant
multi-target designs in our final application of our overall strategy
to a large set of multi-target RNA design instances from a
representative benchmark (Section~\ref{sec:results}).

@@

%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
%
  As we elaborate in subsequent sections, our approach relies on
  breaking down the energy function $\Energy{\R}{S}$ into additive
  components, each depending on only few sequence positions. Given
  $\R$, we express $\Energy{\R}{S}$ as the sum of energy contributions $f(S)$
  over a set $\F$ (of functions $f:\S_n\to\real$),
  s.t.~$\Energy{\R}{S}=\sum_{f\in\F} f(S)$. This captures realistic RNA
  energy models---including nearest neighbor models,
  e.g.~\citep{Turner2009}, and even pseudoknot models,
  e.g. \citep{Andronescu2010}, while bounding the dependencies to
  sequence positions introduced by each single $f$. Formally, define
  the \Def{dependencies} $\dep(f)$ of $f$ as the minimum set of sequence
  positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all
  sequences $S$ and $S'$ that have equal nucleotides at the positions in $\I$.

  Each set $\F$ of functions (on sequences of length $n$) induces a
  \Def{dependency graph} on sequence positions, namely the hypergraph
  $G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$.  Our algorithms
  will critically rely on a \Def{tree decomposition} of the dependency
  graph, which we define below.
\begin{definition}[Tree decomposition and width]
  \label{def:treedecomp}
  Let $G=(X, E)$ be a hypergraph. A \Def{tree decomposition} of $G$ is
  a pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and (for
  each $v\in T$) $\chi(v)\subseteq X$ is a set of vertices assigned to
  the node $v\in T$, such that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item for all $x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
\item for all $e\in E$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}
The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
$\width(T,\chi) = \min_{u\in T} |\chi(u)| - 1 $. The \Def{treewidth}
of $G$ is the smallest width of any tree decomposition of $G$.
\end{definition}

\section{An fixed-parameter tractable (FPT) algorithm for the
  partition function and sampling of Boltzmann-weighted designs}
\label{sec:FPT}

@@

\subsection{Partition function and Boltzmann sampling through stochastic backtrack}\label{sec:PF}
The minimum energy, counting, and partition function
over an RNA design network can be computed by dynamic programming based
on a tree decomposition of the network's dependency graph
(i.e. cluster tree elimination).
%This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.
We focus on the efficient computation of the partition
function. %Observe that counting is a special case of the partition function and can be obtained by setting $\beta =0$.%; moreover, efficient energy minimization can be obtained by a simple $(+,\times)\to (\min,+)$ algebraic substitution and dropping the transformation of energy contributions to Boltzmann weights.

@@

For a set $\Y$ of sequence positions, write $\partseqs(\Y)$ to
denote the set of all partial sequences that determine exactly the positions
in $\Y$; furthermore, given partial sequences $\val$ and
$\val'$, we define the \Def{combined partial sequence $\substitute{\val}{\val'}$} such that
$$
(\substitute{\val}{\val'})_i :=
\begin{cases}
  \val'_i & \text{if } \val'_i\in \Sigma\\
  \val_i & \text{otherwise}
\end{cases}
$$

@@

Let us now consider the computation of the partition function.
Given is the RNA design network $\network=(\X,\F)$ and its cluster
tree decomposition $(T,\chi,\phi)$.  W.l.o.g., we assume that $T$ is
connected and contains a dedicated node $r$, with
$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$, added as a virtual
root $r$ connected to a node in each connected component of $T$.  Now,
we consider the set of directed edges $\edgesToR{}$ of $T$ oriented to
$r$; define $T_r(u)$ as the induced subtree of
$u$. Algorithm~\ref{alg:pf} computes the partition function by passing
messages along these directed edges $u\to v$ (i.e. always from some
child $u$ to its parent $v$). Each message $m$ is a function that depends
on the positions $\dep(m)\subseteq \X$ and yields a partition function
in $\real\cup\{\infty\}$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Supp. Mat.~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function:
\begin{math}
  \partfun{\network} = \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

\begin{algorithm}[t]
  \KwData{Cluster tree $(T,\chi,\phi)$} \KwResult{Messages
    $\Message{u}{v}$ for all $(u\to{}v)\in T$; i.e.~partition
    functions of the subtrees of all $v$ for all possible partial
    sequences determining exactly the positions $\separator{u}{v}$.}
  \For{$u\to{}v\in T$ in postorder}{
    \For{$\val\in\partseqs(\separator{u}{v})$}{
      $x\gets 0$\;
      \For{$\val'\in\partseqs(\difference{u}{v})$}{
        $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
        ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{u'}{u}}{\substitute{\val}{\val'}}$ for $(u'\to{}u)\in T$ )\;
        $x \gets x + p$\;
      }
      $\evalfor{\Message{u}{v}}{\val} \gets x$\;
    }
    \Return {$m$}\;
  }
  \caption{FPT computation of the partition function using dynamic
    programming, i.e.~cluster tree elimination
    \citep[CTE]{Dechter2013}. The postorder traversal guarantees that
    when processing edge $u\to{}v$, all messages
    $\Message{u'}{u}$, corresponding to DP matrices, have been computed before. }\label{alg:pf}
\end{algorithm}

The partition functions can then direct a stochastic backtracking procedure to sample sequences from the Boltzmann distribution based on the given design network $\network$.
%
For an expanded cluster tree, after the messages $\Message{u}{v}$ for
the edges in $\edgesToR{}$ are generated by Algorithm \ref{alg:pf},
one can repeatedly call Algorithm~\ref{alg:sampling}, each time
randomly drawing another sequence from the Boltzmann distribution.

@@

We define the \emph{maximum separator size} $s$ as
$\max_{u,v\in V} | \separator{u}{v} |$ and denote by  $D$ the maximum size of
$\difference{u}{v}$ over $(u,v)\in\edgesToR{}$.  In the absence
of specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and
$\mathcal{O}(|V|\cdot4^s)$ space
(Supp. Mat.~\ref{appsec:algcomplexity}); Alg.~\ref{alg:sampling} would
require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per sample on arbitrary
tree decompositions
(Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

% \begin{algorithm}
%  \KwData{Node $u$, partial sequence $\val\in\partseqs(\separator{u}{v})$;\newline
%  Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}[\val']$, $\forall (u'\to{}v')\in T$ and $\val'\in\partseqs(\separator{u'}{v'})$.}
%  \KwResult{Boltzmann-distributed random partial sequence for the subtree rooted at $u$, specializing a partial sequence $\val$.}
%  \Fn{\Sample$(u,\val)$}{
%    $r \gets \Random(\evalfor{\Message{u}{v}}{\val})$\;
%    \For{$\val'\in\partseqs(\difference{u}{v})$}{

%      $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
%      ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{u'}{u}}{\substitute{\val}{\val'}}$ for $(u'\to{}u)\in T$ )\;
%   	  $r \gets r - p$\;
%   	  \If{$r<0$}{
%   	    $\val_{\rm res} \gets \substitute{\val}{\val'}$\;
%   	  \For{$(u'\to{}u)\in T$}{
%   	      $\val_{\rm res} \gets \substitute{\val_{\rm res}}{\Sample(u',\substitute{\val}{\val'})}$\;
%      }
%   	  \Return {$\val_{\rm res}$}
%   	  \;}
%  }
%  }
%  \caption{Stochastic backtrack algorithm for partial sequences in the Boltzmann distribution.}\label{alg:sampling}
% \end{algorithm}

\begin{algorithm}
 \KwData{Cluster tree $(T,\chi,\phi)$ and partition functions $\Message{u'}{v'}$ for all $(u'\to{}v')\in T$.}
 \KwResult{Boltzmann-distributed random sequence $\val$.}
 $\val$ \gets $\varnothing$\;
 \For{$u\to{}v\in T$ in preorder}{
   $r \gets \text{uniform random number between $0$ and $\evalfor{\Message{u}{v}}{\val}$}$\;
   \For{$\val'\in\partseqs(\difference{u}{v})$}{

     $p \gets$ product( $exp(-\beta \evalfor{f}{\substitute{\val}{\val'}})$ for $f\in \phi(u)$ )\\
     ${}\quad\qquad \cdot\ $product( $\evalfor{\Message{u'}{u}}{\substitute{\val}{\val'}}$ for $(u'\to{}u)\in T$ )\;
     $r \gets r - p$\;
     \If{$r<0$}{
       $\val \gets \substitute{\val}{\val'}$\;
     }
   }
 }
 \Return {$\val$}\;

 \caption{Stochastic backtrack algorithm for partial sequences in the
   Boltzmann distribution. Processing the edges $u\to{}v\in T$ in
   preorder ensures that $\val$ invariantly determines all positions
   of $v$ outside the subtree of $u$.}\label{alg:sampling}
\end{algorithm}

%For example, for design to a single non-crossing target in the nearest neighbor energy model, the complexity of Algs.~\ref{alg:pf} and \ref{alg:sampling} degrades to linear time (as reported for IncaRNAtion), since there are only $O(n)$ many functions and nodes (in the sequence length $n$). Moreover, the dependencies are tree-like owing to the tree-like non-crossing structure; this implies a constantly bounded maximum treewidth.

\begin{theorem}[Complexities]\label{th:complexities}
  For sequence length $n$, $k$ target structures and treewidth $w$
  % and a base pair dependency graph having $c$ connected components%
  , $t$ sequences
  are generated from the Boltzmann distribution  in
  $O( n\, k \, 4^{w+1} + t\, n\, k )$ time.
  %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
  %$\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
\end{theorem}

By this theorem, Boltzmann sampling in our setting is fixed parameter tractable (FPT) in the tree width $w$ (since the complexity is polynomial for fixed $w$).
%
As shown in Supp. Mat.~\ref{sec:improvedComplexity}, the complexity of the precomputation can be further improved to
$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ is the maximum number of connected components represented in a node of the tree decomposition ($c\le w+1$).

\subsection{Sequence features, constraints, and energy
  models.}\label{sec:energy_models}
\todo[inline]{Better motivate why we want different energy functions etc}
%
The functions in $\F$ allow expressing complex features of the
sequences alone, e.g.\ rewarding or penalizing specific sequence
motifs, as well as features depending on the target structures.
Furthermore, constraints, which enforce or forbid features, are
naturally expressed by assigning infinite penalties to invalid
(partial) sequences. Finally, but less obviously, the framework
captures various RNA energy models with bounded dependencies, which we
describe briefly.
